---
title: "The Story of Success"
subtitle: "Unveiling Educational Patterns Across Tunisia"
---

```{r}
#| label: setup
#| include: false

# Set library path and load packages
.libPaths(Sys.getenv("R_LIBS_USER"))
library(tidyverse)
library(knitr)
library(kableExtra)
library(scales)

# Load the data
source("R/load_data.R")

# Calculate key metrics
total_students <- sum(school_data$nb_students, na.rm = TRUE)
total_passed <- sum(school_data$nb_passed, na.rm = TRUE)
total_failed <- sum(school_data$nb_failed, na.rm = TRUE)
overall_rate <- round((total_passed / total_students) * 100, 1)
```

## Chapter 1: The Big Picture {.unnumbered}

::: {.callout-tip appearance="simple"}
## 🎯 The Central Question
**What determines whether a Tunisian student succeeds?** Is it where they live? The level they're studying? Or something else entirely?
:::

Imagine **`r format(total_students, big.mark = ",")`** students across Tunisia, each with dreams and aspirations. Every year, they face the challenge of passing their academic year. 

> **The verdict?** `r format(total_passed, big.mark = ",")` students succeeded, while `r format(total_failed, big.mark = ",")` faced setbacks.

```{r}
#| label: fig-hero
#| fig-cap: "The Tale of Two Outcomes: Success vs. Struggle"
#| fig-width: 10
#| fig-height: 6

# Create dramatic visualization
outcome_data <- data.frame(
  status = c("SUCCESS", "STRUGGLE"),
  count = c(total_passed, total_failed),
  color = c("#27ae60", "#e74c3c")
)

outcome_data <- outcome_data %>%
  mutate(
    pct = round(count / sum(count) * 100, 1),
    label = paste0(format(count, big.mark = ","), "\n(", pct, "%)")
  )

ggplot(outcome_data, aes(x = status, y = count, fill = status)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = label), vjust = -0.3, size = 6, fontface = "bold") +
  scale_fill_manual(values = c("SUCCESS" = "#27ae60", "STRUGGLE" = "#e74c3c")) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "THE VERDICT",
    subtitle = paste0("Overall Success Rate: ", overall_rate, "%"),
    x = "",
    y = "Number of Students"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 28, color = "#2c3e50"),
    plot.subtitle = element_text(hjust = 0.5, size = 18, color = "#7f8c8d"),
    axis.text.x = element_text(size = 18, face = "bold"),
    panel.grid.major.x = element_blank()
  )
```

---

## Chapter 2: A Nation Divided? {.unnumbered}

::: {.callout-warning appearance="simple"}
## 🗺️ The Geography of Success
Does where you study determine your future? Let's explore the regional disparities.
:::

Tunisia's 24 governorates tell vastly different stories. Some regions shine as beacons of academic excellence, while others struggle to keep pace.

```{r}
#| label: fig-regional-story
#| fig-cap: "The Great Divide: Success Rates Across Governorates"
#| fig-width: 14
#| fig-height: 10

gov_stats <- summary_by_governorate(school_data)

# Categorize performance
gov_stats <- gov_stats %>%
  mutate(
    performance = case_when(
      overall_success_rate >= 85 ~ "🌟 Excellence",
      overall_success_rate >= 75 ~ "✅ Good",
      overall_success_rate >= 65 ~ "⚠️ Needs Attention",
      TRUE ~ "🚨 Critical"
    ),
    bar_color = case_when(
      overall_success_rate >= 85 ~ "#27ae60",
      overall_success_rate >= 75 ~ "#3498db",
      overall_success_rate >= 65 ~ "#f39c12",
      TRUE ~ "#e74c3c"
    )
  )

ggplot(gov_stats, aes(x = reorder(governorate, overall_success_rate), y = overall_success_rate)) +
  geom_segment(aes(xend = governorate, yend = 0), color = "gray80", linewidth = 2) +
  geom_point(aes(color = bar_color), size = 8, show.legend = FALSE) +
  geom_text(aes(label = paste0(overall_success_rate, "%")), 
            hjust = -0.5, size = 4, fontface = "bold") +
  geom_hline(yintercept = 75, linetype = "dashed", color = "#e74c3c", linewidth = 1) +
  annotate("text", x = 3, y = 77, label = "← National Target: 75%", 
           color = "#e74c3c", fontface = "italic", size = 4) +
  scale_color_identity() +
  coord_flip() +
  ylim(0, 100) +
  labs(
    title = "REGIONAL PERFORMANCE RANKING",
    subtitle = "Which governorates lead? Which ones need support?",
    x = "",
    y = "Success Rate (%)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 22, color = "#2c3e50"),
    plot.subtitle = element_text(hjust = 0.5, size = 14, color = "#7f8c8d"),
    axis.text.y = element_text(size = 11, face = "bold")
  )
```

### 🏆 The Champions vs 🆘 The Challengers

```{r}
#| label: champions-challengers

# Top 3 and Bottom 3
top_3 <- gov_stats %>% arrange(desc(overall_success_rate)) %>% head(3)
bottom_3 <- gov_stats %>% arrange(overall_success_rate) %>% head(3)

cat("🏆 TOP PERFORMING GOVERNORATES:\n")
for(i in 1:3) {
  cat(paste0("   ", i, ". ", top_3$governorate[i], " - ", top_3$overall_success_rate[i], "%\n"))
}

cat("\n🆘 GOVERNORATES NEEDING SUPPORT:\n")
for(i in 1:3) {
  cat(paste0("   ", i, ". ", bottom_3$governorate[i], " - ", bottom_3$overall_success_rate[i], "%\n"))
}

gap <- max(gov_stats$overall_success_rate) - min(gov_stats$overall_success_rate)
cat(paste0("\n📊 THE GAP: ", round(gap, 1), " percentage points between best and worst performing regions"))
```

::: {.callout-important}
## 💡 Key Insight
A student in **`r top_3$governorate[1]`** has a significantly higher chance of success than one in **`r bottom_3$governorate[1]`**. This `r round(gap, 1)`-point gap represents thousands of students whose outcomes may be influenced by geography alone.
:::

---

## Chapter 3: The Level Effect {.unnumbered}

::: {.callout-note appearance="simple"}
## 📚 Does difficulty increase with level?
We often assume higher levels are harder. But is this true in Tunisia?
:::

```{r}
#| label: fig-level-pyramid
#| fig-cap: "The Educational Ladder: Success at Every Rung"
#| fig-width: 12
#| fig-height: 8

level_stats <- summary_by_level(school_data)

# Create level categories
level_stats <- level_stats %>%
  mutate(
    difficulty = case_when(
      overall_success_rate >= 90 ~ "Achievable",
      overall_success_rate >= 80 ~ "Moderate",
      overall_success_rate >= 70 ~ "Challenging",
      TRUE ~ "Very Challenging"
    )
  )

ggplot(level_stats, aes(x = reorder(level_name, overall_success_rate), y = overall_success_rate)) +
  geom_col(aes(fill = overall_success_rate), show.legend = FALSE, width = 0.7) +
  geom_text(aes(label = paste0(overall_success_rate, "%")), 
            hjust = -0.2, size = 4, fontface = "bold", color = "#2c3e50") +
  scale_fill_gradient2(low = "#e74c3c", mid = "#f39c12", high = "#27ae60", midpoint = 80) +
  coord_flip() +
  ylim(0, 105) +
  labs(
    title = "SUCCESS BY EDUCATIONAL LEVEL",
    subtitle = "Some rungs of the ladder are harder to climb",
    x = "",
    y = "Success Rate (%)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 20, color = "#2c3e50"),
    plot.subtitle = element_text(hjust = 0.5, size = 14, color = "#7f8c8d")
  )
```

---

## Chapter 4: The Hidden Patterns {.unnumbered}

::: {.callout-tip appearance="simple"}
## 🔍 Deep Dive: What the heatmap reveals
Combining region and level tells a more complete story.
:::

```{r}
#| label: fig-heatmap-story
#| fig-cap: "The Full Picture: Region × Level Performance Matrix"
#| fig-width: 14
#| fig-height: 12

# Create level categories
school_data_cat <- school_data %>%
  mutate(
    level_cat = case_when(
      str_detect(level_name, "سابعة") ~ "7th Year",
      str_detect(level_name, "ثامنة") ~ "8th Year",
      str_detect(level_name, "تاسعة") ~ "9th Year",
      str_detect(level_name, "أولى ثانوي") ~ "1st Secondary",
      str_detect(level_name, "ثانية") ~ "2nd Secondary",
      str_detect(level_name, "ثالثة") ~ "3rd Secondary",
      TRUE ~ "Other"
    )
  )

heatmap_data <- school_data_cat %>%
  filter(!is.na(governorate), level_cat != "Other") %>%
  group_by(governorate, level_cat) %>%
  summarise(
    avg_rate = round(mean(success_rate, na.rm = TRUE), 1),
    .groups = "drop"
  )

ggplot(heatmap_data, aes(x = level_cat, y = reorder(governorate, avg_rate), fill = avg_rate)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = avg_rate, 
                color = ifelse(avg_rate > 75, "white", "black")), 
            size = 3.5, fontface = "bold") +
  scale_fill_gradient2(
    low = "#c0392b", 
    mid = "#f39c12", 
    high = "#27ae60", 
    midpoint = 75,
    name = "Success\nRate (%)"
  ) +
  scale_color_identity() +
  labs(
    title = "PERFORMANCE HEATMAP",
    subtitle = "Dark red = struggle | Yellow = moderate | Green = success",
    x = "Educational Level",
    y = "Governorate"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 20, color = "#2c3e50"),
    plot.subtitle = element_text(hjust = 0.5, size = 12, color = "#7f8c8d"),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    legend.position = "right"
  )
```

---

## Chapter 5: The Distribution Story {.unnumbered}

```{r}
#| label: fig-distribution-story
#| fig-cap: "Where Do Most Schools Fall?"
#| fig-width: 12
#| fig-height: 7

mean_rate <- mean(school_data$success_rate, na.rm = TRUE)
median_rate <- median(school_data$success_rate, na.rm = TRUE)

ggplot(school_data, aes(x = success_rate)) +
  geom_histogram(aes(y = after_stat(count)), binwidth = 5, 
                 fill = "#3498db", color = "white", alpha = 0.8) +
  geom_vline(xintercept = mean_rate, color = "#e74c3c", linewidth = 1.5, linetype = "solid") +
  geom_vline(xintercept = 50, color = "#2c3e50", linewidth = 1, linetype = "dashed") +
  annotate("label", x = mean_rate + 8, y = 800, 
           label = paste0("Mean: ", round(mean_rate, 1), "%"), 
           fill = "#e74c3c", color = "white", fontface = "bold", size = 5) +
  annotate("label", x = 50, y = 600, 
           label = "Passing Line: 50%", 
           fill = "#2c3e50", color = "white", fontface = "bold", size = 4) +
  labs(
    title = "THE BELL CURVE OF SUCCESS",
    subtitle = "Most schools cluster around high success rates - a positive sign!",
    x = "Success Rate (%)",
    y = "Number of Schools"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 22, color = "#2c3e50"),
    plot.subtitle = element_text(hjust = 0.5, size = 14, color = "#7f8c8d")
  )
```

::: {.callout-note}
## 📊 What This Tells Us
The distribution is **left-skewed** - meaning most schools achieve HIGH success rates. This is encouraging! However, the long tail to the left shows that some schools are struggling significantly.
:::

---

## Chapter 6: Heroes & Underdogs {.unnumbered}

### 🌟 The Top Performers

These schools achieved near-perfect success rates while educating significant numbers of students:

```{r}
#| label: heroes

top_performers <- school_data %>%
  filter(nb_students >= 50) %>%
  arrange(desc(success_rate), desc(nb_students)) %>%
  head(10) %>%
  select(school_name, governorate, level_name, nb_students, success_rate)

top_performers %>%
  kable(
    col.names = c("🏫 School", "📍 Governorate", "📚 Level", "👥 Students", "✅ Success %"),
    caption = "🏆 Top 10 High-Volume High-Performing Schools"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  row_spec(1:3, bold = TRUE, background = "#d4edda")
```

### 🆘 Schools Needing Support

These schools have room for improvement and could benefit from targeted interventions:

```{r}
#| label: underdogs

struggling <- school_data %>%
  filter(nb_students >= 50) %>%
  arrange(success_rate) %>%
  head(10) %>%
  select(school_name, governorate, level_name, nb_students, success_rate)

struggling %>%
  kable(
    col.names = c("🏫 School", "📍 Governorate", "📚 Level", "👥 Students", "⚠️ Success %"),
    caption = "🆘 Schools That Could Benefit from Additional Support"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  row_spec(1:3, bold = TRUE, background = "#f8d7da")
```

---

## The Story Continues... {.unnumbered}

::: {.callout-tip}
## 📖 Next Chapter
Head to **[Statistical Analysis](statistics.qmd)** to discover the mathematical evidence behind these patterns and test our hypotheses rigorously.
:::
